---
title: "Getting started with Houdini"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with Houdini}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Houdini inserts tables and figures into a Word document. You give it three
things - a Word template with named bookmarks, a folder of RTF files, and a
mapping between them (usually from an excel document) - and it writes a finished document with Word tables at each bookmark.

This vignette builds a small report from scratch: two tables, one figure, and
a filter.

```{r setup, echo = FALSE, message = FALSE}
library(Houdini)

# ---------------------------------------------------------------------------
# Everything in this chunk stands in for files you would already have: a Word
# template exported from your report shell, and a folder of RTF output from
# SAS. It is hidden because none of it is about using Houdini.
# ---------------------------------------------------------------------------

W <- "http://schemas.openxmlformats.org/wordprocessingml/2006/main"

# A minimal but valid .docx containing the given bookmarks, one per paragraph.
make_template <- function(path, bookmarks) {
  paras <- vapply(seq_along(bookmarks), function(i) {
    sprintf(paste0('<w:p><w:bookmarkStart w:id="%d" w:name="%s"/>',
                   '<w:r><w:t>%s</w:t></w:r>',
                   '<w:bookmarkEnd w:id="%d"/></w:p>'),
            i, bookmarks[i], bookmarks[i], i)
  }, character(1))

  doc <- sprintf(
    '<?xml version="1.0" encoding="UTF-8" standalone="yes"?>
     <w:document xmlns:w="%s"><w:body>%s
       <w:sectPr><w:pgSz w:w="11906" w:h="16838"/>
       <w:pgMar w:top="1440" w:right="1440" w:bottom="1440" w:left="1440"/>
       </w:sectPr></w:body></w:document>',
    W, paste(paras, collapse = ""))

  tmp <- tempfile(); dir.create(tmp)
  dir.create(file.path(tmp, "word", "_rels"), recursive = TRUE)
  writeLines(doc, file.path(tmp, "word", "document.xml"))
  writeLines(
    '<?xml version="1.0" encoding="UTF-8" standalone="yes"?>
     <Relationships xmlns="http://schemas.openxmlformats.org/package/2006/relationships"/>',
    file.path(tmp, "word", "_rels", "document.xml.rels"))
  writeLines(
    '<?xml version="1.0" encoding="UTF-8" standalone="yes"?>
     <Types xmlns="http://schemas.openxmlformats.org/package/2006/content-types">
     <Default Extension="rels" ContentType="application/vnd.openxmlformats-package.relationships+xml"/>
     <Override PartName="/word/document.xml" ContentType="application/vnd.openxmlformats-officedocument.wordprocessingml.document.main+xml"/>
     </Types>', file.path(tmp, "[Content_Types].xml"))

  zip::zip(path, list.files(tmp, recursive = TRUE, all.files = TRUE),
           root = tmp, mode = "mirror", include_directories = FALSE)
  unlink(tmp, recursive = TRUE)
  path
}

# One RTF table page. `parameter` populates the page header, which is what
# Houdini reads when you filter by parameter.
rtf_page <- function(header, rows, parameter = NULL, widths) {
  bounds <- cumsum(widths)
  defs <- paste(sprintf("\\cellx%d", bounds), collapse = "")

  row_xml <- function(cells, hdr = FALSE) {
    tag <- if (hdr) "\\trhdr" else ""
    body <- paste(vapply(seq_along(cells), function(i) {
      align <- if (i == 1L) "l" else "c"
      bold <- if (hdr) "\\b" else ""
      sprintf("{\\q%s%s %s\\cell}", align, bold, cells[i])
    }, character(1)), collapse = "")
    sprintf("\\trowd%s%s%s\\row", tag, defs, body)
  }

  head_sec <- if (is.null(parameter)) "" else sprintf(
    "{\\header\\trowd\\cellx8640{\\ql Parameter: %s\\cell}\\row}", parameter)

  paste0("\\sectd\n", head_sec, "\n", row_xml(header, hdr = TRUE), "\n",
         paste(vapply(rows, row_xml, character(1)), collapse = "\n"), "\n\\sect")
}

write_rtf <- function(path, pages) {
  writeLines(paste0("{\\rtf1\\ansi\n", paste(pages, collapse = "\n"), "\n}"),
             path)
  path
}

# A 1x1 PNG is enough to show how figures are handled
png_bytes <- as.raw(c(
  0x89,0x50,0x4e,0x47,0x0d,0x0a,0x1a,0x0a,0x00,0x00,0x00,0x0d,0x49,0x48,0x44,
  0x52,0x00,0x00,0x00,0x01,0x00,0x00,0x00,0x01,0x08,0x02,0x00,0x00,0x00,0x90,
  0x77,0x53,0xde,0x00,0x00,0x00,0x0c,0x49,0x44,0x41,0x54,0x08,0xd7,0x63,0xf8,
  0xcf,0xc0,0x00,0x00,0x03,0x01,0x01,0x00,0x18,0xdd,0x8d,0xb0,0x00,0x00,0x00,
  0x00,0x49,0x45,0x4e,0x44,0xae,0x42,0x60,0x82))

write_image_rtf <- function(path) {
  hex <- paste(sprintf("%02x", as.integer(png_bytes)), collapse = "")
  writeLines(sprintf(
    "{\\rtf1\\ansi\n\\sectd\n{\\pict\\pngblip\\picwgoal5760\\pichgoal3240\n%s}\n\\sect}",
    hex), path)
  path
}

# Pull word/document.xml out of a .docx so the results below can be checked
read_docx_xml <- function(path) {
  tmp <- tempfile(); dir.create(tmp)
  on.exit(unlink(tmp, recursive = TRUE), add = TRUE)
  utils::unzip(path, exdir = tmp)
  paste(readLines(file.path(tmp, "word", "document.xml"), warn = FALSE),
        collapse = "\n")
}

# --- the "study" -----------------------------------------------------------
project <- file.path(tempdir(), "escapology_study")
dir.create(project, showWarnings = FALSE)
rtf_dir <- file.path(project, "rtf")
dir.create(rtf_dir, showWarnings = FALSE)

widths <- c(2880L, 1920L, 1920L, 1920L)

# Table 1: audience reaction, split by trick class
write_rtf(file.path(rtf_dir, "t_reactions.rtf"), rtf_page(
  header = c("Trick Class\\line    Trick name", "Placebo", "Dose A", "Dose B"),
  rows = list(
    c("Illusions", "52", "49", "51"),
    c("    Apparition", "18", "21", "19"),
    c("    Disappearance", "34", "28", "32"),
    c("Escapology", "47", "44", "46"),
    c("    Straight Jacket", "25", "23", "24"),
    c("    Handcuffs", "22", "21", "22")
  ), widths = widths))

# Table 2: two pages, one per measured parameter
write_rtf(file.path(rtf_dir, "t_noise.rtf"), c(
  rtf_page(
    header = c("Statistic", "Placebo", "Dose A", "Dose B"),
    rows = list(c("Mean", "82.4", "88.1", "91.7"),
                c("SD", "6.2", "5.8", "6.9"),
                c("Range", "71-94", "76-99", "78-103")),
    parameter = "Audience Noise (DB)", widths = widths),
  rtf_page(
    header = c("Statistic", "Placebo", "Dose A", "Dose B"),
    rows = list(c("Mean", "3.1", "5.4", "7.8"),
                c("SD", "0.9", "1.2", "1.6"),
                c("Range", "1-6", "2-9", "3-12")),
    parameter = "No. Fainters (n)", widths = widths)
))

write_image_rtf(file.path(rtf_dir, "f_applause.rtf"))

template <- make_template(file.path(project, "report_shell.docx"),
                          c("Table_Reactions", "Table_Noise", "Figure_Applause"))
```

## What you start with

A Word document containing bookmarks where tables or figures should go:

```{r}
names(Houdini:::extract_bookmarks(template))
```

And a folder of RTF files, one per table or figure:

```{r}
list.files(rtf_dir)
```

## Looking at a table first

Before wiring anything up, `houdini_preview()` renders a single RTF so you can
see how Houdini has parsed it. It writes a self-contained HTML page and opens
it; `browse = FALSE` just returns the path.

```{r}
preview <- houdini_preview(file.path(rtf_dir, "t_noise.rtf"), browse = FALSE)
```

The page header tells you what Houdini found:

```{r, echo = FALSE, results = "asis"}
html <- paste(readLines(preview, warn = FALSE), collapse = "\n")
cat(regmatches(html, regexpr('<div class="sub">[^<]*</div>', html)))
```

Two pages, one per parameter, which is what makes the parameter filter below
work. Preview accepts the same filters as a real run, so you can check a
selection before committing it to a config:

```{r}
noise_only <- houdini_preview(
  file.path(rtf_dir, "t_noise.rtf"),
  parameters = "Audience Noise (DB)",
  browse = FALSE
)
```

## Mapping bookmarks to tables

The mapping is done via an Excel document or a data frame with the same columns.
`Bookmark` and `Dataset` are required; the rest are optional filters.

```{r}
config <- data.frame(
  Bookmark   = c("Table_Reactions", "Table_Noise", "Figure_Applause"),
  Dataset    = c("t_reactions.rtf", "t_noise.rtf", "f_applause.rtf"),
  Parameters = c("", "Audience Noise (DB)", ""),
  Levels     = c("Trick Class", "", ""),
  stringsAsFactors = FALSE
)
config
```

Two filters are doing work here:

- `Parameters` keeps only the *Audience Noise (DB)* page of the two-page table.
- `Levels` keeps only top-level rows of the reactions table, dropping the
  indented trick names beneath each class.

Both accept several values separated by semicolons, e.g. `"Week 1; Week 4"`.

## Checking the config before you run

`houdini_validate()` runs every check the app shows in its warnings panel,
without opening the app and without writing anything. It reports the whole list
at once rather than stopping at the first problem, which is what you want in a
scheduled job:

```{r}
problems <- houdini_validate(template, config, rtf_dir, quiet = TRUE)
nrow(problems)
```

A clean config returns no rows. When something is wrong you get one row per
problem, with the config row it came from and a hint:

```{r}
broken <- config
broken$Bookmark[2] <- "Table_Nosie"   # a typo
houdini_validate(template, broken, rtf_dir, quiet = TRUE)[, c("row", "severity", "message")]
```

In a CI step, stop on anything that would produce a wrong document:

```{r, eval = FALSE}
problems <- houdini_validate(template, config, rtf_dir)
if (any(problems$severity == "error")) stop("config is not ready")
```

## Generating the document

```{r}
status <- apparate(template, config, rtf_dir, quiet = TRUE)
output <- sub(".docx$", "_Houdini_Output.docx", template)
file.exists(output)
```

`apparate()` returns a per-row status list rather than stopping at the first
problem so if one table throws an error the rest can still run fine:

```{r}
vapply(status, function(s) if (is.null(s$err)) "ok" else "error", character(1))
```

A failing row carries a structured error with a hint. Note this runs against a
copy of the template: each run writes its output beside its input, so pointing
a second run at the same template would overwrite the document just produced.

```{r}
scratch <- file.path(project, "scratch.docx")
file.copy(template, scratch)

bad <- apparate(scratch,
                data.frame(Bookmark = "No_Such_Bookmark",
                           Dataset  = "t_noise.rtf",
                           stringsAsFactors = FALSE),
                rtf_dir, quiet = TRUE)
conditionMessage(bad[["1"]]$err)
Houdini:::err_hint(bad[["1"]]$err)
```

## Checking what came out

The generated document holds proper Word tables, not images of tables:

```{r}
doc <- xml2::read_xml(read_docx_xml(output))
ns <- c(w = W)
c(tables   = length(xml2::xml_find_all(doc, ".//w:tbl", ns = ns)),
  drawings = length(xml2::xml_find_all(doc, ".//w:drawing", ns = ns)))
```

The filters took effect. Only the requested parameter survived:

```{r}
texts <- xml2::xml_text(xml2::xml_find_all(doc, ".//w:t", ns = ns))
c("Audience Noise page kept" = any(grepl("82.4", texts)),
  "Fainters page dropped"    = !any(grepl("3.1", texts)))
```

And the level filter kept the trick classes while dropping the individual
tricks:

```{r}
c("Illusions kept"   = any(grepl("Illusions", texts)),
  "Apparition dropped" = !any(grepl("Apparition", texts)))
```

## Working from Excel

For anything reproducible, keep the mapping in a excel document. `Export Excel` in the app writes this format for you.

```{r}
sheet <- file.path(project, "config.xlsx")
writexl::write_xlsx(config, sheet)

from_excel <- file.path(project, "from_excel.docx")
file.copy(template, from_excel)

status <- apparate(from_excel, sheet, rtf_dir, quiet = TRUE)
vapply(status, function(s) if (is.null(s$err)) "ok" else "error", character(1))
```

If you want to pass a data frame just pass that rather than the excel file path. When `quiet` is `FALSE`, a log is written next to the output recording every row, the filters applied, and any warnings.

## Using the app

`run_app()` opens the same pipeline as a Shiny app. It is the easier way to
build a configuration in the first place: load the template and RTF folder,
click a row to preview it, and click column headers or rows in the preview to
exclude them.

```{r, eval = FALSE}
run_app()
```

The two workflows share a format, so the usual pattern is to build a
configuration interactively, export it to Excel, then run it can from a script
whenever the RTFs are regenerated or reimported back into the app.

## Regenerating

Re-running over a document that Houdini already produced replaces the injected
content in place rather than adding a second copy, so it is safe to run
repeatedly as the source RTFs change:

```{r}
again <- apparate(output, config, rtf_dir, quiet = TRUE)
final <- sub("\\.docx$", "_Houdini_Output.docx", output)

doc2 <- xml2::read_xml(read_docx_xml(final))
length(xml2::xml_find_all(doc2, ".//w:tbl", ns = c(w = W)))
```

Still three pieces of content, not six.

```{r, include = FALSE}
unlink(project, recursive = TRUE)
unlink(c(preview, noise_only))
```

## Where to go next

- `?apparate` for the full set of run options
- `?houdini_preview` for previewing a single file
- `?parse_xl` for the recognised config columns
